Windowed Developer Topic Analysis

### 2009-04-08 lda-paper.tex
sent to mike and ric

%   At different times, developers focus on different topics and tasks
%   during development.  These different topics and tasks might
%   be obvious to developers familar with the project, but other
%   stakeholders, like managers, might be less aware of the topics of
%   development. Imagine the scenario where developers unexpectedly had to deal with
%   issues orthogonal to the features suggested, at the end of the
%   iteration a manager would wonder where all the effort went to.
%   Currently it is difficult for stakeholders to
%   audit what actually occurred and when certain features or topics
%   were focussed on. We propose a visualization of the
%   changing development topics over time. Our approach uses a topic
%   analysis tool like Latent Dirichlet Allocation (LDA) or Latent
%   Semantic Indexing (LSI). Instead of applying the tool to the
%   entire corpus we apply it to windows (periods) of the corpus. This allows us to
%   see the evolving stream of topics of development occuring
%   over time.

@@

During different periods, developers focus on different topics and
different tasks during development. Stake-holders, such as managers, who are
not directly involved with the project's development might have less awareness of the
project and the topics and tasks that the developers have focussed on.
Imagine the scenario where a manager is trying to discover where all the
developers' effort went after features have not been implemented because the
developers unexpectedly had to deal with issues orthogonal to these
features.  We propose to help stakeholders understand the topics of
development by extracting topics from the descriptions of changes that
programmers provide when they make commits.  We leverage topic analysis
tools such as Latent Dirichlet Allocation (LDA) or Latent Semantic Indexing
(LSI) to analyze periods, such as months, within a project's lifetime,
whereas previous work applied LDA and LSI to to a project's entire lifetime.  These
tools associate descriptions with independent topics extracted from these
descriptions.  
We propose a visualization of changing development topics
over time that allows us to see the evolving stream of topics of development
occurring over time. We demonstrate that windowed topic analysis offers
benefits over entire lifetime topic analysis because of the
independence of topics between windows.

%   In this lab report we seek to explore how topics shift over time in
%   the source control repository of a project. We analyze commit
%   comments, and even source code per revision per each time window and
%   extract the topics prevelant in that window. We expect that topic
%   sets will change over time and the change in topic sets indicate a
%   change in the focus of development. We suppose this could help us to
%   segment the revisions on the time line to discover parts of an
%   iteration. For topic extraction we use \lda (LDA).

@@

% In the presence of fine grained changes and detailed changelogs

@@

% many changelogs how to sort by change type

@@

Commits in SCSs are not necessarily consistent, they can have multiple
purposes, thus we hope that multipurposes manifest themselves as
topics interleaved within the commits.  Sometimes a single revision to
a file contains multiple bug fixes.  Others have demonstrated that by
using powerful topic analysis techniques like Latent Semantic Indexing
(LSI) or Latent Dirichlet Allocation (LDA) one can separate topics
from these mixed comments. These topics often describe separate ideas
that are interleaved throughout the development. A topic represents a
groups of commit comments which are related to each other by their
content.  In this case a topic is a set of tokens that are related to
each other.

@@

Some terms we will be using for the rest of the paper include:
message, word distribution, topics and trends. \emph{Messages} are
blocks of text written by developers. In this paper messages will be
the CVS commit comments made when the user commits changes to files in
a repository. \emph{Word distributions} are the summary of a message
by word count. The word distribution will be over all messages looked
at but since most words will not appear in a message, a word
distribution is effectively a word count divided by the message
size. A \emph{topic} is essentially a set of words which form a word
distribution that is unique and independant within the set of
documents in our total corpus. One could think of a topic as the
center distribution of a group of messages. In this paper we often
summarize topics by their top 10 most frequent words in their word
distribution.  A \emph{trend} is a one or more similar topics which
reoccur over time. Trends are important because they are sets of
topics which reoccur over the development of a project.

@@

In order to infer or associate the expertise of an author with topics
extracted from SCS, Linstead et al proposed an author-sourcecode model
using LDA~\cite{10.1109/MSR.2007.20,NIPS2007637,1321709}. This model
is essentially an extension of the author-topic (AT)
model~\cite{1036902}, which associated authors and topics extracted
with LDA.

@@

In our first exploratory pass we wanted to see if LDA could provide
interesting topics extracted from a real system. We took the
repository of MySQL 3.23, extracted the commits, grouped them by 30
day non-overlapping windows then applied LDA to each of these
windows. We asked LDA to make 20 topics per window and we would pay
attention to the top 10 most frequent words in that topic.  We found
there were common words across topic clusters, such as diffs, annotate
and history, these words probably should be viewed as stop
words. There were notable transitional topics such as in the first
window the word ``bitkeeper'' appears (probably MySQL they adopted
Bitkeeper for their source control system) yet in the following
windows there were no mentions of Bitkeeper. ``RENAME'' also only
appeared once in the first window and never again. Per each topic we
tried to name the purpose of the topic, we found that with little
familarity with the code base that naming the topic was not difficult.

A subsampling of the notable topic words is displayed in Table
\ref{tab:portability}.  To name the topics we picked out terms which
seem to summarize those topics.  After viewing these topic is was
obvious that we would have to combine try to track the evolution of
topics over the windows or rank topics by similarity, thus if they
were similar enough we joined them into a trend. Figure \ref{fig:lda}
displays a sample plot of the extracted topics in table
\ref{tab:portability}.

%It is apparent that we have to compare these topics some how by
%itemset or similarity.

@@

\section{Methodology}
\begin{figure*}
  \centering
  \includegraphics[width=0.9\textwidth]{commit-to-topics} 
  \caption{How commits are analyzed and aggregated into Topics and Trends. Commits are first extracted, then abstracted into word counts or word distributions. These word distributions are then given to topic clusterer like LDA which finds independant word distributions (topics) that these documents are related too and clusters or associates these documents with these topics.}
  \label{fig:commits}
\end{figure*}

Our methodology is to first extract the commits from repositories such
as CVS or Bitkeeper. Then we extract the revisions and commits,
grouping the revisions into commits if necessary. We extract the
commit log message of each commit, count the occurence of each word in
the message, remove stop words, and then produce word distributions
for each message.

@@

We then abstract these topics by comparing them to each other.  Each
topic is compared to each other topic in the system, given a certain
threshold of top 10 similarity, like 8/10 words. Then given this topic
similarity matrix, we find the transitive closure of similar topics,
that is we fill flood along the similarity arcs until we have
partitioned the topics into clusters of similar topics.  These
clusters of topics are called trends. A trend is probably interesting
and valuable if there is more than 1 topic in the trend.

@@

\begin{figure}
  \centering
  \includegraphics[width=0.45\textwidth]{transitiveclosure}
  \caption{Clusting topics by the transitive closure (connectedness or
    a fill flood along arcs) of topic similarity. Nodes are topics and
    arcs imply similarity. Similarity could be a measure such as
    topics share 8 out of 10 top ten words.}
\label{fig:closure}
\end{figure}

@@

We have come up with multiple ways to visualize these topics and
trends.  For all of these plots if we trends which have continuous
segments we plot those segments as joined across time.  We have a
compressed stacked trend view, which displays trends arbitrarily
across the y axis but are placed where they occur on the time-based x
axis. Another plot we have is the trend-time plot, a each trend gets
it own distinct-y value , while the x-axis is time; these topics are
then each plotted on their own line across time as they occur. The
last plot we have is the histogram plot, where we plot each trend on
its own line but stack up the segments of the trend, much like a plot
of counts or a histogram.

The first visualization is a compressed view, we try to sink the
larger trends to the bottom, once the larger trends are stacked, we
fill in the gaps with the smaller trends in the same window and then
stack the smaller trends on top. Although there is a chance that gaps
will be left due to non-optimal stacking in practice there are lot of
small trends (90\% of all trends contain 1 topic) which fill in these
gaps quickly. The trend segments that share the same color are the
same trend, colors are randomly chosen for each trend, so color
similarity is meaningless, only exact color matching matters.
The compressed view is nice because it shows all of the
information and the repeating continuous trends are easy to pick out,
unfortunately repeating discontinuous trends are harder to spot.

The second visualization is the trend-time plot, it displays repeating
trends more clearly by dedicating a horizontal line for trend segments
dedicated to one trend. This means that if a trend contains
discontinuous segments the segments appear on the same line. One down
side to this view is the least common trends need to be pruned
otherwise it'd take too many horizontal lines to plot them.

The 3rd visualization is the histogram plot which is a lot like the
time-trend plot except the trends are plotted together by stacking to
their row so time information is lost and the trends are ordered by
the number of topics in the trend. The histogram plot shows the count
of instances of a trend and thus indicates which trends occur the
most. Unfortunately due to the large number of topics given allotted
space it is often best to crop off the trends with only one topic,
otherwise the tail is long.

@@

\begin{figure}
  \centering
  \includegraphics[width=0.45\textwidth]{class-smear-plot-crop-scaled}
  \caption{Trend Time Line: Trends plotted over time of MaxDB 7.500. Time in months are plotted along the X-Axis, each row on the Y-axis is associated with a trend ranked by size in descending order}         
  \label{fig:trendtimeline}
\end{figure}

@@

%   1. Interesting smears 
%   2.   postgresql
\subsection{Postgresql}
%   3.   mysql

PostgreSQL did not have many trends with two or more topics, but it
did have one large mega-trend which stretched across most of its
existence. Much of the mega-trend consisted of stop words but it did
include topics which had words such as patch, fix, update.

@@

Once stop words were removed the some of the purposes became more
clear. The mega- trend disappeared, the largest trend became bug fixes
submitted by Dal Zotto which is a short trend.

@@

%XXX needs results

@@

>>>>

%   At different times, developers focus on different topics and tasks
%   during development.  These different topics and tasks might
%   be obvious to developers familiar with the project, but other
%   stake-holders, like managers, might be less aware of the topics of
%   development. Imagine the scenario where developers unexpectedly had to deal with
%   issues orthogonal to the features suggested, at the end of the
%   iteration a manager would wonder where all the effort went to.
%   Currently it is difficult for stake-holders to
%   audit what actually occurred and when certain features or topics
%   were focussed on. We propose a visualization of the
%   changing development topics over time. Our approach uses a topic
%   analysis tool like Latent Dirichlet Allocation (LDA) or Latent
%   Semantic Indexing (LSI). Instead of applying the tool to the
%   entire corpus we apply it to windows (periods) of the corpus. This allows us to
%   see the evolving stream of topics of development occurring
%   over time.

@@

During different periods, developers focus on different topics and
different tasks during development. Stake-holders, such as managers, who are
not directly involved with the project's development might have less awareness of the
project and the topics and tasks that the developers have focussed on.
Imagine the scenario where a manager is trying to discover where all the
developers' effort went after features have not been implemented because the
developers unexpectedly had to deal with issues orthogonal to these
features.  We propose to help stake-holders understand the topics of
development by extracting topics from the descriptions of changes that
programmers provide when they make commits.  We leverage topic analysis
tools such as Latent Dirichlet Allocation (LDA) or Latent Semantic Indexing
(LSI) to analyze periods, such as months, within a project's lifetime,
whereas previous work applied LDA and LSI to to a project's entire lifetime.  These
tools associate descriptions with independent topics extracted from these
descriptions.  
We propose a visualization of changing development topics
over time that allows us to see the evolving stream of topics of development
occurring over time. We demonstrate that windowed topic analysis offers
benefits over entire lifetime topic analysis because of the
independence of topics between windows.

%   In this lab report we seek to explore how topics shift over time in
%   the source control repository of a project. We analyze commit
%   comments, and even source code per revision per each time window and
%   extract the topics prevalent in that window. We expect that topic
%   sets will change over time and the change in topic sets indicate a
%   change in the focus of development. We suppose this could help us to
%   segment the revisions on the time line to discover parts of an
%   iteration. For topic extraction we use \lda (LDA).

@@

% In the presence of fine grained changes and detailed change-logs

@@

% many change-logs how to sort by change type

@@

Commits in SCSs are not necessarily consistent, they can have multiple
purposes, thus we hope that these multiple purposes manifest themselves as
topics interleaved within the commits.  Sometimes a single revision to
a file contains multiple bug fixes.  Others have demonstrated that by
using powerful topic analysis techniques like Latent Semantic Indexing
(LSI) or Latent Dirichlet Allocation (LDA) one can separate topics
from these mixed comments. These topics often describe separate ideas
that are interleaved throughout the development. A topic represents a
groups of commit comments which are related to each other by their
content.  In this case a topic is a set of tokens that are related to
each other.

@@

Some terms we will be using for the rest of the paper include:
message, word distribution, topics and trends. \emph{Messages} are
blocks of text written by developers. In this paper messages will be
the CVS commit comments made when the user commits changes to files in
a repository. \emph{Word distributions} are the summary of a message
by word count. The word distribution will be over all messages looked
at but since most words will not appear in a message, a word
distribution is effectively a word count divided by the message
size. A \emph{topic} is essentially a set of words which form a word
distribution that is unique and independent within the set of
documents in our total corpus. One could think of a topic as the
center distribution of a group of messages. In this paper we often
summarize topics by their top 10 most frequent words in their word
distribution.  A \emph{trend} is a one or more similar topics which
reoccur over time. Trends are important because they are sets of
topics which reoccur over the development of a project.

@@

In order to infer or associate the expertise of an author with topics
extracted from SCS, Linstead et al proposed an author-source-code model
using LDA~\cite{10.1109/MSR.2007.20,NIPS2007637,1321709}. This model
is essentially an extension of the author-topic (AT)
model~\cite{1036902}, which associated authors and topics extracted
with LDA.

@@

In our first exploratory pass we wanted to see if LDA could provide
interesting topics extracted from a real system. We took the
repository of MySQL 3.23, extracted the commits, grouped them by 30
day non-overlapping windows then applied LDA to each of these
windows. We asked LDA to make 20 topics per window and we would pay
attention to the top 10 most frequent words in that topic.  We found
there were common words across topic clusters, such as diffs, annotate
and history, these words probably should be viewed as stop
words. There were notable transitional topics such as in the first
window the word ``bitkeeper'' appears (probably MySQL they adopted
Bitkeeper for their source control system) yet in the following
windows there were no mentions of Bitkeeper. ``RENAME'' also only
appeared once in the first window and never again. Per each topic we
tried to name the purpose of the topic, we found that with little
familiarity with the code base that naming the topic was not difficult.

A sub-sampling of the notable topic words is displayed in Table
\ref{tab:portability}.  To name the topics we picked out terms which
seem to summarize those topics.  After viewing these topic is was
obvious that we would have to combine try to track the evolution of
topics over the windows or rank topics by similarity, thus if they
were similar enough we joined them into a trend. Figure \ref{fig:lda}
displays a sample plot of the extracted topics in table
\ref{tab:portability}.

%It is apparent that we have to compare these topics some how by
%item-set or similarity.

@@

\section{Methodology}
\begin{figure*}
  \centering
  \includegraphics[width=0.9\textwidth]{commit-to-topics} 
  \caption{How commits are analyzed and aggregated into Topics and Trends. Commits are first extracted, then abstracted into word counts or word distributions. These word distributions are then given to topic clusterer like LDA which finds independent word distributions (topics) that these documents are related too and clusters or associates these documents with these topics.}
  \label{fig:commits}
\end{figure*}

Our methodology is to first extract the commits from repositories such
as CVS or Bitkeeper. Then we extract the revisions and commits,
grouping the revisions into commits if necessary. We extract the
commit log message of each commit, count the occurrence of each word in
the message, remove stop words, and then produce word distributions
for each message.

@@

We then abstract these topics by comparing them to each other.  Each
topic is compared to each other topic in the system, given a certain
threshold of top 10 similarity, like 8/10 words. Then given this topic
similarity matrix, we find the transitive closure of similar topics,
that is we fill flood along the similarity arcs until we have
partitioned the topics into clusters of similar topics. Figure
\ref{fig:closure} provides an example of clustering topics by the
topic similarity. These clusters of topics are called trends. A trend
is probably interesting and valuable if there is more than 1 topic in
the trend.

@@

\begin{figure}
  \centering
  \includegraphics[width=0.45\textwidth]{transitiveclosure}
  \caption{Topic similarity demonstrated by clustering topics by the transitive closure (connectedness)
    of topic similarity. Nodes are topics and
    arcs imply similarity. Similarity could be a measure such as
    topics share 8 out of 10 top ten words.}
\label{fig:closure}
\end{figure}

@@

We have come up with multiple ways to visualize these topics and
trends.  For all of these plots if we trends which have continuous
segments we plot those segments as joined across time.  We have a
compressed stacked trend view, shown in Figure \ref{fig:topicsmear}
and Figure \ref{fig:zoomedsmear}, which displays trends arbitrarily
across the y axis but are placed where they occur on the time-based x
axis. Another plot we have is the trend-time line plot, shown in
Figure \ref{fig:timeline}, each trend gets it own distinct-y value ,
while the x-axis is time; these topics are then each plotted on their
own line across time as they occur. Our final plot is the histogram
plot, shown in Figure \ref{fig:histogram} where we plot each trend on
its own line but stack up the segments of the trend, much like a plot
of counts or a histogram .

The first visualization is a compressed view (Figure \ref{fig:topicsmear}
and Figure \ref{fig:zoomedsmear}), we try to sink the
larger trends to the bottom, once the larger trends are stacked, we
fill in the gaps with the smaller trends in the same window and then
stack the smaller trends on top. Although there is a chance that gaps
will be left due to non-optimal stacking in practice there are lot of
small trends (90\% of all trends contain 1 topic) which fill in these
gaps quickly. The trend segments that share the same color are the
same trend, colors are randomly chosen for each trend, so color
similarity is meaningless, only exact color matching matters.
The compressed view is nice because it shows all of the
information and the repeating continuous trends are easy to pick out,
unfortunately repeating discontinuous trends are harder to spot.

The second visualization is the trend-time line plot (Figure
\ref{fig:timeline}), it displays repeating trends more clearly by
dedicating a horizontal line for trend segments dedicated to one
trend. This means that if a trend contains discontinuous segments the
segments appear on the same line. One down side to this view is the
least common trends need to be pruned otherwise it'd take too many
horizontal lines to plot them.

The third visualization is the histogram plot (Figure
\ref{fig:histogram}) which is a lot like the time-trend plot except
the trends are plotted together by stacking to their row so time
information is lost and the trends are ordered by the number of topics
in the trend. The histogram plot shows the count of instances of a
trend and thus indicates which trends occur the most. Unfortunately
due to the large number of topics given allotted space it is often
best to crop off the trends with only one topic, otherwise the tail is
long.

@@

\begin{figure}
  \centering
  \includegraphics[width=0.45\textwidth]{class-smear-plot-crop-scaled}
  \caption{Trend-time Line: Trends plotted over time of MaxDB 7.500. Time in months are plotted along the X-Axis, each row on the Y-axis is associated with a trend ranked by size in descending order}         
  \label{fig:trendtimeline}
\end{figure}

@@

%   1. Interesting smears 
%   2.   postgresql

%\subsection{MySQL}
%XXX DO MYSQL?

\subsection{Postgresql}
%   3.   mysql

%XXX Redo?

PostgreSQL did not have many trends with two or more topics, but it
did have one large trend which stretched across most of its
existence. Much of the large trend consisted of stop words but it did
include topics which had words such as patch, fix, update.

@@

Once stop words were removed the some of the purposes became more
clear. The large trend disappeared, the largest trend became bug fixes
submitted by Dal Zotto which is a short trend.

@@

%XXX REDO?

@@

%XXX REDO?
